Follow-up Samples

### 2017-02-12 submission/manuscript_outline_20161024.tex
incorporate updated SRA and corrected RF runs into manuscript

\textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
health problem with early detection being used as a key component in
mitigating deaths due to the disease. Previous research suggests a link
between fecal bacterial microbiome and CRC. The overall objective of our
study was to investigate the changes in the bacterial microbiome after
surgery in patients with lesion (i.e.~adenoma or carcinoma).
Specifically, we wanted to identify what within the community was
different within those undergoing surgical removal of lesion. We also
wanted to investigate the use of the bacterial microbiome and Fecal
Immunoglobulin Test (FIT) to build models which could classify
individuals as having a lesion or as before surgery.\\
\textbf{Results:} Adenoma individual's bacterial microbiome were more
similar to their pre-surgery sample then those with carcinoma (P-value =
0.00198). Their change in FIT was also significantly different then
those with carcinoma (P-value = 2.15e-05). There was no significant
difference in any indivdiual OTU between samples before and after
surgery (P-value \textgreater{} 0.125). A model with a total of 37
variables was able to classify lesion (AUC = 0.847 - 0.791) while the
model to classify samples as before surgery had 33 variables (AUC = 0.79
- 0.651). There was a significant decrease in the model positive
probability for the after surgery sample being either a lesion or before
surgery sample (P-value = 1.91e-11 and 6.72e-12). In total there were 14
OTUs that were common to both models and the majority of these
classified to commensal bacteria (Bacteroides, Blautia, Streptococcus,
and Clostridiales).\\
\textbf{Conclusions:} Our data suggests that treatment not only
significantly reduces the probability of having a colonic lesion within
our models but also causes detectable changes in the bacterial
microbiome. Further surveillance of these individuals will enable us to
determine whether models such as the ones we present could be used to
predict recurrence of colorectal cancer.

@@

Colorectal cancer (CRC) continues to be a leading cause of cancer
related deaths and is the second most common cancer death among men aged
40-79 years of age {[}1,2{]}. Over the last few years death due to the
disease has seen a significant decrease, thanks mainly to improvements
in screening {[}1{]}. However, despite this improvement there are still
approximately 50,000 deaths from the disease a year {[}2{]}. It is
estimated that around 5-10\% of all CRCs can be explained by autosomal
dominant inheritance {[}3{]}. The vast majority of CRCs are not
inherited and the exact etiology of the disease has not been well worked
out {[}2{]}. Although many risk factors have been identified {[}2{]} and
non-invasive screening techniques have started to be put into consistent
use {[}4,5{]} there has been an additional increase in the incidence of
CRC in the younger population.

This increased incidence of CRC in the younger population is concerning
since having either an adenoma or carcinoma increases ones risk for
future adenomas or carcinomas {[}6--8{]}. This increased risk can also
carry with it an increased chance of mortality due to this recurrence
{[}9,10{]}. Therefore, there has been a great amount of interest in
early risk stratification tools {[}11,12{]} that can help identify those
that may be at most susceptibility to reccurence. Concurrently, there
has also been a lot of interest in new areas that could have a role in
disease pathogenesis, such as the gut bacterial microbiome.

There has been promising work on the bacterial microbiome and it's
ability to be able to complement existing screening methods such as
Fecal Immunoglobulin Test (FIT) or act alone as a screening tool
{[}13,14{]}. There has also been research into how this microbiome could
be altered directly on tumor tissue itself {[}15{]}. A few studies have
now even shown how this microbiome {[}16{]} or specific members within
it {[}17{]} could be directly involved with the pathogenesis of CRC.
These studies have helped to provide a tantilzing link between the
bacterial microbiome and CRC. However, at this present time there
remains limited information on the bacterial microbiome before and after
successful surgery for removal of the adenoma or carcinoma and whether
it changes at all.

In this study we investigated what happened to the bacterial microbiome
before and after surgery for indivdiuals with either adenoma or
carcinoma. Our anlaysis includes both alpha and beta diversity analysis
along with investigation of individual operational taxonomic units
(OTUs). We also utilized Random Forest models and observed how these
models as well as specific OTUs within them performed before (initial)
and after (follow up) surgery. We also used these models to look for
similar important OTUs to identify the crucial OTUs for not only
classifying initial and follow up samples but also lesion or normal.

@@

\textbf{\emph{Bacterial Community and FIT}} Based on the thetayc
distance metric, comparing the initial to the follow up samples, there
was a significant difference between the adenoma and carcinoma groups
(P-value = 0.00198) {[}Figure 1a{]}. There was also a significant
difference in change in FIT between initial and follow up samples
(P-value = 2.15e-05) {[}Figure 1b{]}. The whole community structure
before and after surgery was visualized on NMDS graphs for both adenoma
{[}Figure1c{]} (PERMANOVA = 0.997) and carcinoma {[}Figure 1d{]}
(PERMANOVA = 0.002). When all initial and follow up samples were
compared to each other there was no significant overall difference
between them (PERMANOVA = 0.085). There was no significant difference
between initial and follow up samples for observed OTUs, Shannon
diversity, or evenness after correction for multiple comparisons
{[}Table S1{]}. There was also no significant difference between initial
and follow up samples for any single OTU {[}Figure S1{]}.

\textbf{\emph{Previously Associated Cancer Bacteria}} We next examined
whether there were differences in previously well described carcinoma
associated OTUs. These included the OTUs that aligned with
\emph{Porphyromonas asaccharolytica} (Otu000153), \emph{Fusobacterium
nucleatum} (Otu000226), \emph{Parvimonas micra} (Otu000460), and
\emph{Peptostreptococcus stomatis} (Otu000653). First, the carcinoma
samples showed a significant difference between initial and follow up
samples for \emph{Peptostreptococcus stomatis} (P-value = 0.00913),
\emph{Parvimonas micra} (P-value = 0.0185), and \emph{Porphyromonas
asaccharolytica} (P-value = 0.00913) whereas there were no signficant
differences in any of these OTUs in the adenoma samples. Second, when
these OTUs were present, there was a clear magnitude difference based on
whether they were from adenoma or carcinoma {[}Figure 2{]}. However,
only a small percentage of those with adenoma or carcinoma were positive
for any of these OTUs.

\textbf{\emph{Full and Reduced Model}} Since differences were observed
between initial and follow up samples and only a small number of
individuals were positive for previously associated CRC bacteria; we
next investigated if we could create models that could adequately
classify and adjust either lesion or before sample probability based on
the bacterial community and FIT. The lesion model had an AUC range of
0.723 to 0.795 versus the before sample model which had an AUC range of
0.451 to 0.67 after 100 iterations of 20 repeated 10-fold cross
validations. Interestingly, identification of the most important
variables and reducing the models to only these factors increased the
AUC in the lesion model (0.791 - 0.847) and before sample model (0.651 -
0.79).

The test set AUC range for the full and reduced lesion model were
similar to that reported for the training set AUC ranges and the ROC
curve ranges overlap each other {[}Figure 3a{]}. The ROC curve for the
final lesion model used was within the range of both the full and
reduced lesion model {[}Figure 3a{]}. Interestingly, the test set AUC
range for the before sample model performed much better then the
training set AUCs. Both the full and reduced before sample models
overlapped with each other {[}Figure 3b{]} there was a marked decrease
in the ROC curve for the final model used.

\textbf{\emph{Most Important Variables}} The reduced models were built
based on the most important variables to the respective full model. For
the lesion model there were a total of 37 variables {[}Figure S2{]}
whereas for the before sample model there were a total of 33 variables
{[}Figure S3{]}. For both models FIT resulted in the largest decrease in
MDA {[}Figure S2a \& S3a{]}.

\textbf{\emph{Positive Probability after Lesion Removal}} Regardless of
model used there was a significant decrease in the positive probability
of either the sample being lesion or a before sample on follow up
{[}Figure 4 \& S4{]} (full lesion P-value = 1.11e-11, reduced lesion
P-value = 1.91e-11, initial and follow up P-value = 6.71e-12, and
reduced initial and follow up P-value = 6.72e-12).

For the full and reduced lesion model there was a significant difference
in the classification for the lesion model between predicted and actual
(P-value = 4.19e-10 and 6.98e-10, respectively) but not for the before
sample model (P-value = 1.00 and 1.00). However, the lesion model
correctly kept the one individual who still had a carcinoma on follow up
above the cut off threshold {[}Figure 4a \& S4a{]} for a positive call
while the before sample models did not {[}Figure 4b \& S4b{]}.

\textbf{\emph{Common OTUs to both Models}} There were a total of 14 OTUs
that were common to both models. Of these OTUs the most common taxonomic
identifications were to Blautia, Bacteroides, Streptococcus, and
Clostridiales. The majority of these OTUs had classifications to
bacteria typically thought of as commensal {[}Table S2{]}.

\textbf{\emph{Treatment and Time Differences}} There was no difference
in the amount of change in positive probability for either the full or
reduced lesion model for either chemotherapy (P-value = 0.821 and 0.821)
or radiation therapy (P-value = 0.69 and 0.981). Although the before
sample model was similar there was a significant decrease in positive
probability for those treated with chemotherapy (P-value = 7.04e-04 and
5.07e-03). There was a significant difference in time elapsed in the
collection of the follow up sample between adenoma and carcinoma samples
(uncorrected P-value = 7.59e-05) with time passed being lower for
adenoma (253 +/- 41.3 days) then carcinoma (351 +/- 102 days).

@@

There was no difference in alpha diversity metrics between the initial
and follow up samples {[}Table S1{]}. Yet based on thetayc there was a
significant difference between inital and follow up samples {[}Figure
1a{]} with carcinoma samples being more dissimilar to their initial
samples then adenoma samples. The change in FIT was also different
between initial and follow up for those with adenoma or carcinoma
{[}Figure 1b{]}. There was also a significant difference in the overall
bacterial community structure for those in the carcinoma group {[}Figure
1d{]} before and after surgery but not for the adenoma group {[}Figure
1c{]}. Investigation of the OTU relative abundance before and after
surgery found no single OTU to be significantly different {[}Figure
S1{]}.

Interestingly, when only previously associated cancer bacteria were
investigated only 3 out of the 4 had a significant decrease in relative
abundance between initial and follow up for the carcinoma group and 0
out of the 4 were significant for the adenoma group. This data suggests
that the changes that may be important in the bacterial microbiome after
surgery are not necessarily any one specific bacterium, a clear
depression of CRC related bacteria, or any addition of new or depressed
bacterium. Instead it may be due to a shift of the bacterial community
as a whole between existing members of the host's microbiome.

We next created a model that incorporated FIT and the bacterial
microbiome to either be able to classify lesions (adenoma or carcinoma)
or before samples in order to confirm that the community was what was
responding or changing due to surgery. We found that the OTUs that made
up the most important variables to the model overwhelmingly belonged to
comensal bacteria. With only the lesion model having a single OTU from a
previously associated cancer bacterium (\emph{Porphyromonas
asaccharolytica}). Using only these important OTUs and FIT both models
(lesion and before sample) had a significantly decreased positive
probability of either lesion or being a before sample on follow up
{[}Figure 4 \& S4{]}. Further confirmation of the importance of the
changes of comensal bacteria to these classifications was that a total
of 14 OTUs were common to both models and the vast majority belonged to
regular residents of our gut community.

There was no difference for the majority of models tested for
differences in positive probablity based on whether chemotherapy or
radiation was received. There was also no difference in the thetayc
distance metric based on length of time between initial and follow up
sample between adenoma and carcinoma. These results would indicate that
the findings described were specific to the surgical intervention and
that differences observed between carcinoma and adenoma samples can not
be simply attributed to collection time between samples.

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool {[}13,14{]}
by exploring what happens to the bacterial community after surgical
removal of a lesion. Based on previous work by Arthur, et al. {[}18{]}
it may not be surprising to have E.coli as one of the most important
OTUs and one that was common to both models. Interestingly, many of the
most important OTUs had taxonomic identification for resident gut
microbes. This could suggest that the bacterial community is one of the
first components that could change during the pathogenesis of disease.
These bacterial microbiome changes could be the first step in allowing
more inflammatory bacterium to gain a foothold within the colon
{[}19{]}.

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are a number of reasons why this may
have occured. First, one potential explanation is that even with surgery
and a shift of the bacterial community these specific bacteria still
persisted within the colon. Second, the bacteria even if they were
reduced in relative abundance they were still present within the gut.
Third, is that they were not present in enough individuals to be able to
classify those with and without disease with a high degree of accuracy.
Fourth, is that it is possible that our Random Forest models were able
to get the same information from measures such as FIT or other OTUs.
Finally, it is also possible that all of these potential explanations
could have played a role. Regardless, our observations would suggest
that an individual's resident bacteria have a large role to play in
disease initiation and could change in a way that allows predictive
models to lower the positive probability of a lesion after surgery
{[}Figure 4{]}. It should be noted that our study does not argue against
the importance of these CRC associated bacteria in the pathogenesis of
disease but rather that the models do not utilize these specific
bacteria for classification purposes (lesion or before sample). In fact,
it is possible that these CRC associated bacteria are important in the
transition from adenoma to carcinoma and would be one explanation as to
why in our data we not only see high initial relative abundances, in
certain individuals, but also large decreases in relative abundance in
those with carcinoma but not in those with adenoma after surgery
{[}Figure 2{]}.

One limitation of our study is that we do not know whether individuals
who were still classified as positive by the lesion model eventually had
a subsequent CRC diagnosis. This information would help to strengthen
the case for our Random Forest based model to have kept a number of
individuals above the cutoff threshold even though at follow up they
were diagnosed as no longer having a lesion. Another limitation is that
we do not know if adding modern tests such as the stool DNA test
{[}20{]} could help improve our overall AUC. Another limitation is that
this study drew heavily from those with caucasian ancestry. The results
may not be immediately representative of those with either Asian or
African ancestry. Finally, although our training and test set are
relatively large we still run the risk of overfitting or having a model
that may not be immediately extrapolateable to other populations. We've
done our best to safeguard against this by not only running 10-fold
cross validation but also having over 100 different 80/20 splits to try
and mimic the type of variation that might be expected to occur. The
time difference in collection of sample between adenoma and carcinoma
that was observed could have affected our observed results. However, our
random forest based model, specifically the lesion trained model, were
still able to adjust positive probabilities for both adenoma and
carcinoma to similar amounts despite this.

Another interesting outcome was that within figure 3 the before sample
model showed better test AUC results then the training set AUC. This may
have occured because the training AUC that was determined from 20
repeated 10 fold cross validation removed samples at random and did not
take into account that they were matched samples. Another potential
reason is that the model itself may be overfit since the total number of
samples was not that large. However, the lesion model did not suffer
from these discrepencies and similar conclusions can be drawn solely
from this model. Regardless, further independent studies will need to be
carried out to verify our findings since not only are we dealing with
feces, which could be very different than the communities present on the
actual tissue, but also are dealing with correlations that may not be
representative of the true pathogensis of disease.

@@

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al
{[}13{]}. In brief, study exclusion involved those who had already
undergone surgery, radiation, or chemotherapy, had colorectal cancer
before a baseline fecal sample could be obtained, had IBD, a known
hereditary non-polyposis colorectal cancer, or Familial adenomatous
polyposis. Samples used to build the models for prediction were
collected either prior to a colonoscopy or between 1 - 2 weeks after.
The bacterial microbiome has been shown to normalize within this time
period {[}21{]}. Our follow up data set had a total of 67 individuals
that not only had a sample as described but also a follow up sample
between 188 - 546 days after surgery and treatment had been completed.
This study was approved by the University of Michigan Institutional
Review Board. All study participants provided informed consent and the
study itself conformed to the guidelines set out by the Helsinki
Declaration.

\textbf{\emph{FIT and 16S rRNA Gene Sequencing}} FIT was analyzed as
previously published using both OC FIT-CHEK and OC-Auto Micro 80
automated system (Polymedco Inc.) {[}22{]}. 16S rRNA gene sequencing was
completed as previously described by Kozich, et al. {[}23{]}. In brief,
DNA extraction used the 96 well Soil DNA isolation kit (MO BIO
Laboratories) and an epMotion 5075 automated pipetting system
(Eppendorf). The V4 variable region was amplified and the resulting
product was split between three sequencing runs with normal, adenoma,
and carcinoma evenly represented on each run. Each group was randomly
assigned to avoid biases based on sample collection location.

\textbf{\emph{Sequence Processing}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described {[}23{]}. The general processing workflow
using mothur was as follows: Paired-end reads were first merged into
contigs, quality filtered, aligned to the SILVA database, screened for
chimeras, classified with a naive Bayesian classifier using the
Ribosomal Database Project (RDP), and clustered into Operational
Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
neighbor clustering algorithm. The numer of sequences for each sample
was rarified to 10521 in an attempt to minimize uneven sampling.

\textbf{\emph{Lesion Model Creation}} The Random Forest {[}24{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main testing and training of the model
completed on a data set of 490 individuals. This model was then applied
to our follow up data set of 67 individuals. The model included data on
FIT and the bacterial microbiome. Non-binary data was checked for near
zero variance and OTUs that had near zero variance were removed. This
pre-processing was performed with the R package caret (v6.0.73).
Optimization of the mtry hyperparameter involved taking the samples and
making 100 different 80/20 (train/test) splits of the data where normal
and lesion were represented in the same proportion within both the whole
data set and the 80/20 split. Each of these splits were then run through
20 repeated 10-fold cross validations to optimize the mtry
hyperparameter by maximizing the AUC (Area Under the Curve of the
Receiver Operator Characteristic). This resulting model was then tested
on the 20\% of the data that was originally held out from this overall
process. Once the ideal mtry was found the entire 490 sample set was
used to create the final Random Forest model on which classifications on
the 67-person cohort was completed. The default cutoff of 0.5 was used
as the threshold to classify individuals as positive or negative for
lesion. The hyperparameter, mtry, defines the number of variables to
investigate at each split before a new division of the data is created
with the Random Forest model.

>>>>

\textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
health problem with previous research suggesting that a link may exist
between the fecal bacterial microbiome and CRC. The overall objective of
our study was to test the hypothesis that changes in the bacterial
microbiome occur after surgery in patients with lesions (i.e.~adenoma or
carcinoma). Specifically, we wanted to identify what within the
community was different before and after surgical removal of said
lesion.

\textbf{Results:} The bacterial microbiome in the pre and post surgery
samples of the individuals with adenoma were similar while those with
carcinoma were not (P-value = 0.000472). There was no difference in the
relative abundance of any OTU between the pre and post surgery samples
(P-value \textgreater{} 0.151). A model with a total of 53 variables was
able to classify lesion (AUC = 0.811 - 0.866) while a model built to
classify samples as before surgery had 70 variables (AUC = 0.641 -
0.805). The post surgery sample for both models had a decrease in the
positive probability for either lesion or before surgery sample (P-value
= 3.28e-04 and 6.71e-12). In total there were 23 OTUs that were common
to both models and the majority of these classified to commensal
bacteria (Bacteroides, Clostridiales, Blautia, and Ruminococcaceae).\\
\textbf{Conclusions:} Our data supports the hypothesis that there are
differences in the bacterial microbiome between pre and post surgical
samples. With individuals with carcinoma having more drastic changes to
the overall community then those with adenoma. However, changes to
commensal bacteria were common to individuals with either adenoma or
carcinoma suggesting that these bacteria may be central to initial polyp
formation and transition to carcinoma.

@@

Colorectal cancer (CRC) continues to be a leading cause of cancer
related deaths and is the second most common cancer death among men aged
40-79 years of age {[}1,2{]}. Over the last few years death due to the
disease has seen a significant decrease, thanks mainly to improvements
in screening {[}1{]}. However, despite this improvement there are still
approximately 50,000 deaths from the disease a year {[}2{]}.

Recently, there has been promising work on the bacterial microbiome and
it's ability to be able to complement existing screening methods such as
Fecal Immunoglobulin Test (FIT) or act alone as a screening tool
{[}3,4{]}. There has also been research into how this microbiome could
be altered directly on tumor tissue itself {[}5{]}. A few studies have
also shown how this microbiome {[}6{]} or specific members within it
{[}7{]} could be directly involved with the pathogenesis of CRC. These
studies have helped to provide a tantalizing link between the bacterial
microbiome and CRC. Yet there remains limited information on the
bacterial microbiome before and after surgery for removal of lesion
(adenoma or carcinoma) and whether the community changes at all.

In this study we tested the hypothesis of whether that the bacterial
microbiome changes after surgery for individuals with a lesion. Our
analysis included both alpha and beta diversity analysis along with
investigation of individual operational taxonomic units (OTUs). We also
utilized Random Forest models and observed how these models as well as
specific OTUs within them performed pre (initial) and post (follow up)
surgery. We also used these models to look for similar important OTUs to
identify the crucial OTUs for not only classifying initial and follow up
samples but also lesion or normal.

@@

\textbf{\emph{Bacterial Community and FIT}} We first wanted to test
whether there were any broad differences between initial and follow up
samples based on lesion being either adenoma or carcinoma. What we found
was that the bacterial community in those with carcinoma were more
dissimilar (as measured by thetayc) to their initial sample then those
with adenoma (P-value = 0.000472) {[}Figure 1a{]}. We also found that
there were larger changes in fecal blood (measured by FIT) for those
with carcinoma versus adenoma (P-value = 2.15e-05) {[}Figure 1b{]}. The
broad shift in bacterial community structure before and after surgery
was visualized using NMDS for both adenoma {[}Figure1c{]} (PERMANOVA =
0.999) and carcinoma {[}Figure 1d{]} (PERMANOVA = 0.005). Interestingly,
when initial and follow up samples were compared to each other,
regardless of whether they were adenoma or carcinoma (lesion), there was
no significant overall difference between them (PERMANOVA = 0.09). When
investigating more broad alpha diversity metrics there was no difference
found between initial and follow up samples for lesion, adenoma, or
carcinoma for any metric tested {[}Table S1{]}. We also observed that
there was no difference in OTU relative abundance between initial and
follow up samples for lesion, adenoma, or carcinoma {[}Figure S1{]}.

\textbf{\emph{Previously Associated Cancer Bacteria}} Previous
literature has suggested that a number of oral microbes may be important
in CRC pathogenesis {[}8{]}. So we next examined whether there were
differences in previously well described carcinoma associated OTUs.
These included the OTUs that aligned with \emph{Porphyromonas
asaccharolytica} (Otu000202), \emph{Fusobacterium nucleatum}
(Otu000442), \emph{Parvimonas micra} (Otu001273), and
\emph{Peptostreptococcus stomatis} (Otu001682). There was a difference
in relative abundance in initial and follow up samples for lesion and
carcinoma for \emph{Parvimonas micra} (P-value = 0.0116), and
\emph{Porphyromonas asaccharolytica} (P-value = 0.00842). In contrast,
there was no difference in relative abundance in any of these OTUs for
those with adenoma {[}Figure 2{]}. We also observed that only a small
percentage of those with adenoma or carcinoma were positive or had an
appreciable relative abundance of these respective OTUs {[}Figure 2{]}.

\textbf{\emph{Full and Reduced Model}} We next wanted to identify if
there were any common bacterial microbiome changes in individuals with
adenoma or carcinoma. In order to investigate this we created two
different models: one to classify lesion versus normal and one to
classify pre versus post samples based on the bacterial community and
FIT. The lesion model had an AUC range of 0.73 to 0.797 while the
initial sample model had an AUC range of 0.485 to 0.686 after 100
iterations of 20 repeated 10-fold cross validations. By identifying the
most important variables for each respective model and then reducing
them to only these factors we were able to increase the AUC in the
lesion model (0.811 - 0.866) and initial sample model (0.641 - 0.805).\\
The test set AUC range for the full and reduced lesion model were
similar to that reported for the training set AUC ranges and the ROC
curve ranges overlap with each other {[}Figure 3a{]}. The ROC curve for
the final lesion model used was within the range of both the full and
reduced lesion model {[}Figure 3a{]}. Interestingly, the test set AUC
range for the initial sample model performed much better then the
training set AUCs. Both the full and reduced initial sample models
overlapped with each other {[}Figure 3b{]} but there was a marked
decrease in the ROC curve for the final before sample model used.

\textbf{\emph{Common OTUs to both Models}} The reduced models were built
based on the most important variables to either classification model.
For the lesion model there were a total of 53 variables {[}Figure S2{]}
whereas for the initial sample model there were a total of 70 variables
{[}Figure S3{]}. For both models FIT resulted in the largest decrease in
MDA {[}Figure S2a \& S3a{]}. There were a total of Bacteroides,
Clostridiales, Blautia, and Ruminococcaceae. The majority of these OTUs
had classifications to bacteria typically thought of as commensal
{[}Table S2{]}.

\textbf{\emph{Positive Probability after Lesion Removal}} If there were
common OTUs for individuals with adenoma and carcinoma that were
different versus normal controls we would expect to find a decrease in
the positive probability of the follow up sample to be either lesion or
a before sample. This is what we observed regardless of model used
(lesion or before sample) or whether it was built on the full or reduced
variable data set {[}Figure 4 \& S4{]} (full lesion P-value = 1.62e-04,
reduced lesion P-value = 3.28e-04, initial and follow up P-value =
6.71e-12, and reduced initial and follow up P-value = 6.71e-12).

Although there was a decrease in positive probability for all models
there were differences on whether the models classified the samples as
still having lesion. This was most evident for the full and reduced
lesion model where there was a large difference between predicted and
actual (P-value = 4.88e-01 and 4.7e-01, respectively). However, the
before sample model did not have this disparity between predicted and
actual (P-value = 1.00 and 1.00). Even though the lesion model varied
greatly between predicted and actual it was able to correctly keep the
one individual who still had a carcinoma on follow up above the cut off
threshold {[}Figure 4a \& S4a{]} for a positive call while the before
sample models did not {[}Figure 4b \& S4b{]}.

\textbf{\emph{Treatment and Time Differences}} After observing these
changes in positive probability we wanted to assess whether additional
treatments, such as chemotherapy and radiation, could have an impact on
the results that we observed. There was no difference in the amount of
change in positive probability for either the full or reduced lesion
model for either chemotherapy (P-value = 0.184 and 0.184) or radiation
therapy (P-value = 0.942 and 0.839). Although the before sample model
was similar there was a significant difference in decreased positive
probability for those treated with chemotherapy (P-value = 1.27e-03 and
6.18e-04). Finally, we wanted to know if the length of time between the
initial and follow up sample could be a possible con-founder. Within our
study there was a significant difference in the time elapsed in the
collection of the follow up sample between adenoma and carcinoma samples
(uncorrected P-value = 7.59e-05) with time passed being lower for
adenoma (253 +/- 41.3 days) then carcinoma (351 +/- 102 days).

@@

From our results there were some large observed differences in the
bacterial microbiome between pre and post surgery samples based on
whether the individual had an adenoma or carcinoma. There was much
larger difference between initial and follow up samples based on the
thetayc distance metric and in fecal blood as measured by FIT for
individuals with carcinoma versus adenoma {[}Figure 1{]}. However, there
were no differences between initial and follow up samples for Shannon
Diversity, observed OTUs, or evenness regardless of whether the
individual had an adenoma or carcinoma {[}Table S1{]}. There was also no
differences in relative abundance of any specific OTU for lesion,
adenoma, or carcinoma {[}Figure s1{]}.

Although there were no differences when investigating all OTUs, when
looking specifically at four OTUs that taxonomically classified to
previously suggested cancer causing microbes we found that only 3/4 had
a decrease in relative abundance between initial and follow up for those
with carcinoma and 0/4 had differences for those with adenoma. This data
would suggest that these specific OTUs may be important in the
transition of an adenoma to a carcinoma but less so in the initiation of
an adenoma from benign tissue.

We next created a model that incorporated FIT and the bacterial
microbiome to either be able to classify lesions (adenoma or carcinoma)
or initial samples in order to find common OTUs in the community that
change for both adenoma and carcinoma. What we found was that the
commonly associated CRC bacteria were not higly represented within our
models but rather that OTUs that made up the most important variables
overwhelmingly belonged to commensal bacteria. With only the lesion
model having a single OTU from a previously associated cancer bacterium
(\emph{Porphyromonas asaccharolytica}). Using only these important OTUs
and FIT both models (lesion and before sample) significantly decreased
positive probability of either lesion or being an initial sample on
follow up {[}Figure 4 \& S4{]}. Further confirmation of the importance
of the changes of commensal bacteria to these classifications was that a
total of 23 OTUs were common to both models and the vast majority
belonged to regular residents of our gut community.

\emph{Problematic paragraph need to amend}

There was no difference for the majority of models tested for
differences in positive probability based on whether chemotherapy or
radiation was received. There was a difference in the length of time
between initial and follow up sample between adenoma and carcinoma.
These results would indicate that the findings described were specific
to the surgical intervention and that differences observed between
carcinoma and adenoma samples can not be simply attributed to collection
time between samples.

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool {[}3,4{]} by
exploring what happens to the bacterial community after surgical removal
of a lesion. Based on previous work by Arthur, et al. {[}9{]} it may not
be surprising to have E.coli as one of the most important OTUs and one
that was common to both models. Interestingly, many of the most
important OTUs had taxonomic identification for resident gut microbes.
This could suggest that the bacterial community is one of the first
components that could change during the pathogenesis of disease. These
bacterial microbiome changes could be the first step in allowing more
inflammatory bacterium to gain a foothold within the colon {[}8{]}.

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are a number of reasons why this may
have occurred. First, one potential explanation is that even with
surgery and a shift of the bacterial community these specific bacteria
still persisted within the colon. Second, the bacteria even if they were
reduced in relative abundance they were still present within the gut.
Third, is that they were not present in enough individuals to be able to
classify those with and without disease with a high degree of accuracy.
Fourth, is that it is possible that our Random Forest models were able
to get the same information from measures such as FIT or other OTUs.
Finally, it is also possible that all of these potential explanations
could have played a role. Regardless, our observations would suggest
that an individual's resident bacteria have a large role to play in
disease initiation and could change in a way that allows predictive
models to lower the positive probability of a lesion after surgery
{[}Figure 4{]}. It should be noted that our study does not argue against
the importance of these CRC associated bacteria in the pathogenesis of
disease but rather that the models do not utilize these specific
bacteria for classification purposes (lesion or before sample). In fact,
it is possible that these CRC associated bacteria are important in the
transition from adenoma to carcinoma and would be one explanation as to
why in our data we not only see high initial relative abundances, in
certain individuals, but also large decreases in relative abundance in
those with carcinoma but not in those with adenoma after surgery
{[}Figure 2{]}.

One limitation of our study is that we do not know whether individuals
who were still classified as positive by the lesion model eventually had
a subsequent CRC diagnosis. This information would help to strengthen
the case for our Random Forest based model to have kept a number of
individuals above the cutoff threshold even though at follow up they
were diagnosed as no longer having a lesion. Another limitation is that
we do not know if adding modern tests such as the stool DNA test
{[}10{]} could help improve our overall AUC. Another limitation is that
this study drew heavily from those with Caucasian ancestry. The results
may not be immediately representative of those with either Asian or
African ancestry. Finally, although our training and test set are
relatively large we still run the risk of over-fitting or having a model
that may not be immediately extrapolate-able to other populations. We've
done our best to safeguard against this by not only running 10-fold
cross validation but also having over 100 different 80/20 splits to try
and mimic the type of variation that might be expected to occur. The
time difference in collection of sample between adenoma and carcinoma
that was observed could have affected our observed results. However, our
random forest based model, specifically the lesion trained model, were
still able to adjust positive probabilities for both adenoma and
carcinoma to similar amounts despite this.

Another interesting outcome was that within figure 3 the before sample
model showed better test AUC results then the training set AUC. This may
have occurred because the training AUC that was determined from 20
repeated 10 fold cross validation removed samples at random and did not
take into account that they were matched samples. Another potential
reason is that the model itself may be over-fit since the total number
of samples was not that large. However, the lesion model did not suffer
from these discrepancies and similar conclusions can be drawn solely
from this model. Regardless, further independent studies will need to be
carried out to verify our findings since not only are we dealing with
feces, which could be very different than the communities present on the
actual tissue, but also are dealing with correlations that may not be
representative of the true pathogenesis of disease.

@@

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al
{[}3{]}. In brief, study exclusion involved those who had already
undergone surgery, radiation, or chemotherapy, had colorectal cancer
before a baseline fecal sample could be obtained, had IBD, a known
hereditary non-polyposis colorectal cancer, or Familial adenomatous
polyposis. Samples used to build the models for prediction were
collected either prior to a colonoscopy or between 1 - 2 weeks after.
The bacterial microbiome has been shown to normalize within this time
period {[}11{]}. Our follow up data set had a total of 67 individuals
that not only had a sample as described but also a follow up sample
between 188 - 546 days after surgery and treatment had been completed.
This study was approved by the University of Michigan Institutional
Review Board. All study participants provided informed consent and the
study itself conformed to the guidelines set out by the Helsinki
Declaration.

\textbf{\emph{FIT and 16S rRNA Gene Sequencing}} FIT was analyzed as
previously published using both OC FIT-CHEK and OC-Auto Micro 80
automated system (Polymedco Inc.) {[}12{]}. 16S rRNA gene sequencing was
completed as previously described by Kozich, et al. {[}13{]}. In brief,
DNA extraction used the 96 well Soil DNA isolation kit (MO BIO
Laboratories) and an epMotion 5075 automated pipetting system
(Eppendorf). The V4 variable region was amplified and the resulting
product was split between three sequencing runs with normal, adenoma,
and carcinoma evenly represented on each run. Each group was randomly
assigned to avoid biases based on sample collection location.

\textbf{\emph{Sequence Processing}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described {[}13{]}. The general processing workflow
using mothur was as follows: Paired-end reads were first merged into
contigs, quality filtered, aligned to the SILVA database, screened for
chimeras, classified with a naive Bayesian classifier using the
Ribosomal Database Project (RDP), and clustered into Operational
Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
neighbor clustering algorithm. The number of sequences for each sample
was rarefied to 10523 in an attempt to minimize uneven sampling.

\textbf{\emph{Lesion Model Creation}} The Random Forest {[}14{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main testing and training of the model
completed on a data set of 490 individuals. This model was then applied
to our follow up data set of 67 individuals. The model included data on
FIT and the bacterial microbiome. Non-binary data was checked for near
zero variance and OTUs that had near zero variance were removed. This
pre-processing was performed with the R package caret (v6.0.73).
Optimization of the mtry hyper-parameter involved taking the samples and
making 100 different 80/20 (train/test) splits of the data where normal
and lesion were represented in the same proportion within both the whole
data set and the 80/20 split. Each of these splits were then run through
20 repeated 10-fold cross validations to optimize the mtry
hyper-parameter by maximizing the AUC (Area Under the Curve of the
Receiver Operator Characteristic). This resulting model was then tested
on the 20\% of the data that was originally held out from this overall
process. Once the ideal mtry was found the entire 490 sample set was
used to create the final Random Forest model on which classifications on
the 67-person cohort was completed. The default cutoff of 0.5 was used
as the threshold to classify individuals as positive or negative for
lesion. The hyper-parameter, mtry, defines the number of variables to
investigate at each split before a new division of the data is created
with the Random Forest model.
